Method to infer relatedness using low coverage ancient DNA

### 2022-02-18 paper_overleaf/main.tex

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic1.png}
    \centering
    \caption{Overview of KIn. A. In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On left, the parents have unrelated chromosomes, and on right one parent has homozygous pair of chromosomes. Bottom row shows the expected proportion of differences along the chromosome in both the cases. B. Overall schematic of the method. Blue boxes show the data files, while green boxes represent scripts.}
    \label{fig0:schematic}
\end{figure}

@@

\subsection{Relatedness classes}
This model is run for all the different cases of relatedness using the corresponding transition matrix, and the likelihood for all the relatedness models is compared. The relatedness for the maximum likelihood model is reported, and the confidence is given by the log likelihood ratio between the two highest likelihood models.  
Hidden state of each window (IBD state) is estimated using standard Viterbi algorithm, and the IBD states corresponding to the most likely model are returned.

@@

\paragraph{ Simulations for estimation of transition matrix}
We use simulations both for estimating the transition matrices for all our models, and for testing and validating our algorithm. All simulations are performed using \textt{msprime} \cite{kelleher_efficient_2016}, followed by simulations of related individuals using a predetermined pedigree (Figure S1).

@@

For artificially mating two individuals, we simulated a recombined chromosome from each individual to be passed on to the progeny. For simulating a recombined chromosome from a diploid individual, we drew the number of recombination points from a Poisson distribution with parameter $rL$ as the recombination rate, and used a uniform distribution to sample the positions of recombination points. We estimated transition matrices corresponding to different cases of relatedness by counting transition between IBD states for a pair of  individuals at a particular relatedness level. We realize that we have low power to detect $4^{th}$ and $5{th}$ degree and cna differentiate these from unrelated. In our final output, we show classifications of $4_{th}$ and $5_{th}$ degrees as unrelated.

\paragraph{Model Comparison}
model with maximum likelihood is reported, and log likelihood ratio is calculated with the two highest likelihood models.

#### Simulations for model evaluation
Apart from the related and unrelated individuals in pop1, we simulated more haploids in three other populations to create scenarios with ascertainment and contamination. We simulated two haploids to form an individual each from two other populations (pop2, pop3) with split time of 3500 and 4500 generations with pop1, and sampling time of 4000 generations and 2000 generations ago respectively. We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from pop1. We test out the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a markov chain using transition matrix shown below:

@@

\section{Supplementary Information}
Beta binomial distribution is a better fit compared to Binomial distribution for proportion of differences in windows.

>>>>

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings with and without ROH. In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On left, the parents have unrelated chromosomes, and on right one parent has homozygous pair of chromosomes. Bottom row shows the expected proportion of differences along the chromosome in both the cases.}
    \label{fig0:schematic}
\end{figure}

@@

\subsection{Relatedness classes}
This model is run for different cases of relatedness including (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{rd}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, Identical) using the corresponding transition matrix, and the likelihood for all the relatedness models is compared. We realized that we have low power to detect $4^{th}$ and $5^{th}$ degree and we can not always differentiate these from unrelated. In our final output, we show classifications of $4^{th}$ and $5^{th}$ degrees as unrelated. Similarly, we observe that we have low power to differentiate between half-siblings, avuncular and grandparent-grandchild, and we merge these classifications as $2^{nd}$ Degree in the final results. The relatedness for the maximum likelihood model is reported, and the confidence is given by the log likelihood ratio between the two highest likelihood models.  
Hidden IBD state of each window ($Z_w$) is estimated using standard Viterbi algorithm, and $\BZ$ corresponding to the most likely model are returned.

@@

We use simulations both for estimating the transition matrices for all our models, and for testing and validating our algorithm. All simulations are performed using \textt{msprime} \cite{kelleher_efficient_2016}, followed by simulations of related individuals using a predetermined pedigree (Figure S1).

\paragraph{ Simulations for estimation of transition matrix}

@@

For artificially mating two individuals, we simulated a recombined chromosome from each individual to be passed on to the progeny. For simulating a recombined chromosome from a diploid individual, we drew the number of recombination points from a Poisson distribution with parameter $rL$ as the recombination rate, and used a uniform distribution to sample the positions of recombination points. We estimated transition matrices corresponding to different cases of relatedness by counting transition between IBD states for a pair of  individuals at a particular relatedness level. We compared the transition matrices for grandparent-grandchild and siblings estimated in this way to a theoretical calculation, and verified that they are very close. However, it is difficult to theoretically calculate transition matrices for other cases like third degree, and so for these cases we use the transition matrices obtained from simulations.

\paragraph{Model Comparison}
Model with maximum likelihood is reported, and log likelihood ratio is calculated with the two highest likelihood models.

Apart from the related and unrelated individuals in pop1, we simulated more haploids in three other populations to create scenarios with ascertainment and contamination. We simulated two haploids to form an individual each from two other populations (pop2, pop3) with split time of 3500 and 4500 generations with pop1, and sampling time of 4000 generations and 2000 generations ago respectively. We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from pop1. We test out the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a markov chain using transition matrix shown below:

@@

\section{Supplementary Information}

Pipeline of KIn
\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sup.png}
    \centering
    \caption{Overall schematic of the method. Blue boxes show the data files, while green boxes represent scripts.}
    \label{figS0:schematic}
\end{figure}

Beta binomial distribution is a better fit compared to Binomial distribution for proportion of differences in windows.
